### 2020-06-23 latex/vnhcs.tex
Finished Hamiltonian (first draft) and started simply actuated systems.

\chapter{Development of Virtual Nonholonomic Constraints}
\section{Preliminaries on Analytical Mechanics}
In classical mechanics, a mechanical system is comprised of \(N\) point masses
representing the center of mass of a physical body, where the position of each
mass is described using three cartesian coordinates (one for each spatial axis). 
The system as a whole is represented by a vector in \(\R^{3N}\), and the
dynamics of the system are computed by deriving the \(3N\) \textit{equations of
motion} (EOM) produced by Newton's second law \(F = m a\).  
While this method works for simple cases, it unfortunately proves impossible
to apply to a majority of mechanical systems since most of the forces
are not known explicitly.

@@

Each generalized coordinate \(q_i\) takes values in \(\Rt{T_i}\), where
\(T_i = \infty\) if \(q_i\) represents a length or \(T_i = 2\pi\) if \(q_i\)
represents an angle.
It is convention to write these as a \textit{configuration} vector
\(q = (q_1,\ldots,q_n) \in \mathcal{Q}\) where the \textit{configuration space}
\(\mathcal{Q}\) has the following structure:
\[
    \mathcal{Q} = \Rt{T_1} \times \cdot \times \Rt{T_n}
\] 
The derivative \(\dot{q} = (\dot{q}_1,\ldots,\dot{q}_n)\) of a configuration
is called a \textit{generalized velocity}, and the joint vector
\((q,\dot{q})\) is called a \textit{state} of the system.

@@

Lagrangian mechanics uses the kinetic energy \(T(q,\dot{q})\) and potential
energy \(P(q)\) of the system to define the Lagrangian
(\ref{eqn:lagrangian_general})
\begin{equation}\label{eqn:lagrangian_general}
    \mathcal{L}(q,\dot{q}) = T(q,\dot{q}) - P(q)
\end{equation}
In actuated mechanical systems, the EOM are described by \(n\) second-order
differential equations obtained from the \textit{Euler-Lagrange
equations} with \textit{input torques} \(\tau \in \R^m\) 
(\ref{eqn:el_eqns_general}). 
\begin{equation}\label{eqn:el_eqns_general}
    \diff{}{t}\left\{ \pdiff{\mathcal{L}}{\dot{q}_i} \right\}
    - \pdiff{\mathcal{L}}{q_i} = B_i(q) \tau
\end{equation}
Here, \(B_i: \mathcal{Q} \rightarrow \R^{1\times m}\) is an input vector 
and the matrix  \(B: \mathcal{Q} \rightarrow \R^{n \times m}\) given by
\[
    B(q) =\begin{bmatrix}
        - & B_1(q) & - \\
          & \vdots & \\
        - & B_n(q) & -
    \end{bmatrix}
\]
is the \textit{input matrix} for the system. If \(m < n\), we say the system is
\textit{underactuated} with degree of underactuation \(n - m\).

Many actuated mechanical systems have quadratic kinetic energies, so that the
Lagrangian can be written explicitly as
\[
    \mathcal{L}(q,\dot{q}) = \dot{q}\tpose D(q) \dot{q} - P(q)
\]
where the \textit{inertia matrix} \(D: \mathcal{Q} \rightarrow \R^{n\times n}\) 
is symmetric, positive definite matrix for all \(q \in \mathcal{Q}\) and the
potential function \(P : \mathcal{Q} \rightarrow \R\) is smooth. 
If this is the case, the Euler-Lagrange equations
(\ref{eqn:el_eqns_general}) reduce to (\ref{eqn:el_eqns}),
\begin{equation}\label{eqn:el_eqns}
    D(q)\ddot{q} + C(q,\dot{q})\dot{q} + \nabla P(q) = B(q)\tau
\end{equation}
where the \textit{Coriolis matrix} \(C(q,\dot{q})\) is of the form
\[
    [C]_{i,j} = \frac{1}{2}\sum\limits_{k = 1}^n 
    \left(\pdiff{[D]_{i,j}}{q_k}  +
     \pdiff{[D]_{i,k}}{q_j} -
     \pdiff{[D]_{k,j}}{q_i}\right)\dot{q}_k
\]

% ---------- Hamiltonian Mechanics ---------- %
\subsection{Hamiltonian Mechanics}
% TODO: Describe how to get the Hamiltonian from the lagrangian, and show that
% you can write the hamiltonian directly in certain conditions of the system

\section{Canonical Control Coordinates for Hamiltonian Systems}
% TODO: Describe the change of coordinates to get into q_u/q_a mode and show
% that the actuator directly affects pa but not pu. Use M for simply actuated
% inertia, D for normal coordinates

>>>>

\chapter{Development of Virtual Nonholonomic Constraints}
\section{Preliminaries on Analytical Mechanics}
In classical mechanics, a mechanical system is comprised of \(N\) point masses.
Each point mass represents the center of mass of a physical body, where the
position of each mass is described using three cartesian coordinates (one for
each spatial axis).  The system as a whole is represented by a vector in
\(\R^{3N}\), and the dynamics of the system are computed by deriving the \(3N\)
\textit{equations of motion} (EOM) produced by Newton's second law \(F = m a\).
While this method works for simple cases, it unfortunately proves impossible to
apply to a majority of mechanical systems since most of the forces are not known
explicitly.

@@

Each generalized coordinate \(q_i\) takes values in \(\Rt{T_i}\), where
\(T_i = \infty\) if \(q_i\) represents a length or \(T_i = 2\pi\) if \(q_i\)
represents an angle.
It is convention to write these as a \textit{configuration} vector 
\(q = (q_1,\ldots,q_n) \in \mathcal{Q}\) 
where the \textit{configuration manifold} \(\mathcal{Q}\) of the system has the
following structure:
\[
    \mathcal{Q} = \Rt{T_1} \times \cdot \times \Rt{T_n}
\] 
The derivative \(\dot{q} = (\dot{q}_1,\ldots,\dot{q}_n)\) of a configuration
is called a \textit{generalized velocity}, and the combined vector
\((q,\dot{q})\) is called a \textit{state} of the system.

@@

Lagrangian mechanics uses the kinetic energy \(T(q,\dot{q})\) and potential
energy \(P(q)\) of the system to define the Lagrangian
(\ref{eqn:lagrangian-general})
\begin{equation}\label{eqn:lagrangian-general}
    \mathcal{L}(q,\dot{q}) = T(q,\dot{q}) - P(q)
\end{equation}
In actuated mechanical systems, the EOM are described by \(n\) second-order
ordinary differential equations (ODEs) obtained from the \textit{Euler-Lagrange
equations} with \textit{generalized input forces} \(\tau \in \R^m\) 
(\ref{eqn:el-eqns-general}). 
\begin{equation}\label{eqn:el-eqns-general}
    \diff{}{t}\left\{ \pdiff{\mathcal{L}}{\dot{q}_i} \right\}
    - \pdiff{\mathcal{L}}{q_i} = B_i(q) \tau
\end{equation}
Here, \(B_i: \mathcal{Q} \rightarrow \R^{1\times m}\)
and the matrix  \(B: \mathcal{Q} \rightarrow \R^{n \times m}\) given by
\[
    B(q) =\begin{bmatrix}
        - & B_1(q) & - \\
          & \vdots & \\
        - & B_n(q) & -
    \end{bmatrix}
\]
is the \textit{input matrix} for the system. If \(m < n\), we say the system is
\textit{underactuated} with degree of underactuation \(n - m\).

Many actuated mechanical systems have quadratic kinetic energies, so that the
Lagrangian can be written explicitly as
\begin{equation}\label{eqn:lagrangian}
    \mathcal{L}(q,\dot{q}) = \frac{1}{2} \dot{q}\tpose D(q) \dot{q} - P(q)
\end{equation}
where the \textit{inertia matrix} \(D: \mathcal{Q} \rightarrow \R^{n\times n}\) 
is a symmetric, positive definite matrix for all \(q \in \mathcal{Q}\) and the
potential function \(P : \mathcal{Q} \rightarrow \R\) is smooth. 
If this is the case, the Euler-Lagrange equations
(\ref{eqn:el-eqns-general}) reduce to (\ref{eqn:el-eqns}),
\begin{equation}\label{eqn:el-eqns}
    D(q)\ddot{q} + C(q,\dot{q})\dot{q} + \nabla P(q) = B(q)\tau
\end{equation}
where the \textit{Coriolis matrix} \(C(q,\dot{q})\) is of the form
\[
    [C]_{i,j} = \frac{1}{2}\sum\limits_{k = 1}^n 
    \left(\pdiff{[D]_{i,j}}{q_k}  +
     \pdiff{[D]_{i,k}}{q_j} -
     \pdiff{[D]_{k,j}}{q_i}\right)\dot{q}_k
\]

% ---------- Hamiltonian Mechanics ---------- %
\subsection{Hamiltonian Mechanics}
% TODO: Describe how to get the Hamiltonian from the lagrangian, and show that
% you can write the hamiltonian directly in certain conditions of the system
Hamiltonian mechanics provides an equivalent representation of the EOM
by converting the \(n\) second-order ODEs generated by Lagrangian mechanics into
\(2n\) first-order ODEs.

To do this, we first define the \textit{conjugate of momentum to \(q_i\)} by
\begin{equation}\label{eqn:p-i}
    p_i(q,\dot{q}) = \pdiff{\mathcal{L}}{\dot{q}_i}(q,\dot{q})
\end{equation}
To ease notation, we will write \(p = (p_1, \ldots, p_n)\) and say that \(p\) is
the \textit{conjugate of momenta to \(q\)} Note that each \(p_i\) is a linear
function of \(\dot{q}\), so one can typically solve for \(\dot{q}(q,p)\) by
inverting the expressions from (\ref{eqn:p-i}). The combined vector \((q,p)\) is
called the \textit{phase} of the system.

Next we define the \textit{Hamiltonian} (\ref{eqn:hamiltonian-general}) 
as the Legendre transform of the Lagrangian \textbf{SOURCE}:
\begin{equation}\label{eqn:hamiltonian-general}
    \mathcal{H}(q,p) = p\tpose \dot{q}(q,p) - \mathcal{L}(q,\dot{q}(q,p))
\end{equation}
The EOM in this framework can be shown to be the \(2n\)
first-order equations
\begin{equation}\label{eqn:hamiltonian-eom-general}
    \begin{cases}
        \dot{q} = \pdiff{\mathcal{H}}{p} \\
        \dot{p} = -\pdiff{\mathcal{H}}{q} + B(q)\tau \\
    \end{cases}
\end{equation}
where \(B(q) \in \R^{n\times m}\) is the input matrix and 
\(\tau \in \R^m\) is the vector of generalized input forces.

If the Lagrangian is given as in (\ref{eqn:lagrangian}), the conjugate of
momenta is computed explicitly by
\[
    p = D(q)\dot{q}
\]
which gives the Hamiltonian (\ref{eqn:hamiltonian})
\begin{equation}\label{eqn:hamiltonian}
    \mathcal{H}(q,p) = \frac{1}{2} p\tpose D^{-1}(q) p + P(q)
\end{equation}
and the EOM from (\ref{eqn:hamiltonian-eom-general}) reduce to 
(\ref{eqn:hamiltonian-eom}).
\begin{equation}\label{eqn:hamiltonian-eom}
    \begin{cases}
        \dot{q} = D^{-1}(q)p \\
        \dot{p_i} = -\frac{1}{2} p\tpose \pdiff{D^{1}}{q_i}(q) p -
        \pdiff{P}{q_i}(q) + B_i(q) \tau \\
    \end{cases}
\end{equation}

\section{Simply Actuated Hamiltonian Systems}
% TODO: Describe the change of coordinates to get into q_u/q_a mode and show
% that the actuator directly affects pa but not pu. Use M for simply actuated
% inertia, D for normal coordinates
Given a mechanical system with phase coordinates \((q,p)\) and Hamiltonian
(\ref{eqn:hamiltonian}), it is not in general obvious how the input forces
\(\tau\) are affecting the conjugate of momenta since they are scaled by
\(B(q)\), which may be quite complicated. Even if \(B(q)\) is constant, multiple
forces may affect each \(p_i\). This section will demonstrate a useful change of
coordinates which results in a simpler input matrix.

Throughout the rest of this thesis, we will make the following assumptions.
\begin{assm}\label{assm:B-const}
    The input matrix \(B : \mathcal{Q} \rightarrow \R^{n\times k}\) is constant
    and full rank.
\end{assm}
\begin{assm}\label{assm:B-orthogonal}
    The input matrix is left-semi-orthogonal, whereby \(B\tpose B = \Id{k}\).
\end{assm}
\begin{assm}\label{assm:B-perp}
    There exists a right-semi-orthogonal matrix 
    \(B^\perp \in \R^{(n-k)\times n}\) for \(B\)
    which is a left-annihilator for \(B\). That is,
    \(B^\perp(B^\perp)\tpose = \Id{(n-k)}\) and
    \(B^\perp B = \Zmat{(n-k) \times k}\) 
\end{assm}

Observe that Assumption \ref{assm:B-orthogonal} can be taken without loss of
generality (WLOG). If \(B\) is a constant full-rank matrix, 
it has a singular-value decomposition 
\(B = U \Sigma V\tpose\) where \(U^{-1} = U\tpose \in \R^{n \times n}\), 
\(V^{-1} = V\tpose \in \R^{n \times k}\), and \(\Sigma \in \R^{n \times k}\) is
defined by
\[
    \Sigma = \begin{bmatrix}
        \sigma_1 & 0 & \cdots & 0 & 0 \\
        0 & \sigma_2 & \cdots & 0 & 0 \\
        \vdots & & \ddots &  & \vdots \\
        0 & 0 & \cdots & 0 & \sigma_k \\
          & - & \Zmat{(n-k)\times k} & -  &\\
    \end{bmatrix}
\]
where \(\sigma_i \neq 0\).
\textbf{SOURCE}
Defining \(T \in \R^{k \times k}\) by
\[
    T = \begin{bmatrix}
        \frac{1}{\sigma_1} & 0 & \cdots & 0 \\
        0 & \frac{1}{\sigma_2} & \cdots & 0 \\
    \vdots & & \ddots & \vdots \\
    0 & 0 & \cdots & \frac{1}{\sigma_k} \\
    \end{bmatrix}
\]
and assigning the input forces to \(\tau = V T \hat{\tau}\), we get a new input
matrix for \(\hat{\tau}\) given by \(\hat{B} = B V T = U \Sigma T\) 
which is still constant and full-rank. In particular, 
\(\hat{B}\tpose \hat{B} = T\tpose \Sigma\tpose \Sigma\tpose T = \Id{k}\).

If \(k = (n-1)\) then if there is a left annihilator \(A\) for
\(B\), then \(B^\perp = A/\norm{A}\) satisfies Assumption
\ref{assm:B-perp}.

To perform our coordinate transformation, we define a transformation matrix
\[
    \mathbf{B} = 
    \begin{bmatrix}
        B^\perp \\
        B\tpose \\
    \end{bmatrix}
\]
Since both \(B^\perp\) and \(B\tpose\) are right-semi-orthogonal, 
\(\mathbf{B}\) is orthogonal. That is,
\(\mathbf{B}\mathbf{B}\tpose = \Id{n}\) which means
\(\mathbf{B}^{-1} = \mathbf{B}\tpose\).

\begin{defn}
    We call \(q_u = B^\perp q\) the \textbf{unactuated coordinates} and 
    \(q_a = B\tpose\) the \textbf{actuated coordinates}, which are obtained
    through the transformation
    \[
        \begin{bmatrix} q_u \\ q_a \end{bmatrix} = \mathbf{B}q
    \]
\end{defn}
